MyTitle

### 2024-10-29 paper/main.tex
updated

@@

\subsubsection{Our Approach}

@@

\chapter{Set-Theoretic Preliminaries}
In this chapter, we introduce the concepts of set theory used in the formalization of this study.
We use first-order logic with the language of set theory, which consists only of only two relation symbols $\in$ and $=$. 
Formulas involving other mathmatical operators that may appear are considered abbreviations for formulas in this language.
Parentheses in formulas are omitted where no confusion arises.
Unless otherwise stated, "a statement holds" means "the statement holds in ZF".

\section{ZF Set Theory and the Axiom of Choice}
\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\text{"} f \text{ is a function on } x \text{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF + AC is denoted by ZFC.
Next, we introduce the well-ordering theorem, as we treat AC in this form.

@@

\section{Forcing}
Forcing is a tequnique used in proving relative consistency and Independence.
We introduce basic concepts of forcing in the context of the transitive countable model (c.t.m.) approach.
In this approach, the relative consistency proof is achieved by using forcing to construct a extended model by adding new sets to an assumed c.t.m.
Let $M$ be a c.t.m. of ZF and $(\Pbb, \leq_{\Pbb})$ be a notion of forcing, which is a pre-ordered set in $M$ with a maximum element $1_{\Pbb}$.

\begin{dfn}
  We define $M^{\Pbb}$, the set of \emph{$\Pbb$-names}, by transfinite recursion on ordinals:
  \begin{enumerate}
    \item $M^{\Pbb}_0 = \emptyset$
    \item $M^{\Pbb}_{\alpha + 1} = \mathcal{P}^M(M^{\Pbb}_{\alpha} \times \Pbb)$
    \item $M^{\Pbb}_{\alpha} = \bigcup_{\beta < \alpha} M^{\Pbb}_{\beta}$ for a limit ordinal $\alpha$
    \item $M^{\Pbb} = \bigcup_{\alpha \in \text{Ord}} M^{\Pbb}_{\alpha}$
  \end{enumerate}
\end{dfn}
Where $\mathcal{P}^M$ denotes the power set operation in $M$. 
We often write a $\Pbb$-name with a dot, e.g., $\dot{x}$.

@@

\begin{dfn} 
  Let $G$ be a generic filter on $\Pbb$ and $\dot{x} \in M^{\Pbb}$. We define the \emph{interpretation} of $\dot{x}$ denoted by $\dot{x}_G$ recursively, 
  $\dot{x}_G = \{\dot{y}_G \mid \exists p \in G (\langle \dot{y}, p \rangle \in \dot{x})\}$. 
\end{dfn}

@@

By choosing $\Pbb$ appropriately, we can construct $M[G]$ with various properties.
What holds or does not hold in $M[G]$ can be identified using the forcing relation.

\begin{dfn} [Forcing relation]
  ATODE
\end{dfn}

@@

@@

\begin{proof}
We outline the proof of this theorem as follows. 
For every $n \in \omega$, let $a_n$ be the following real number:
$$a_n = \{ m \in \omega \mid \exists p \in G (p(n, m) = 1) \}$$
Since $a_n$ are pairwise distinct, $A = \{ a_n \mid n \in \omega \}$ is an infinity set.
$A$ and every $a_n$ are elements of $\Ncal$.
$A$ serves as a counterexample to the well-ordering theorem in $\Ncal$.
Suppose for contradiction that $A$ is well-ordered in $\Ncal$, there exists a injection $f$ from $\omega$ to $A$ in $\Ncal$.
Let $\varphi(g,x,y)$ be a formula that represents the relation $g(x) = y$. 
For every $n \in \omega$, There exists $i \in \omega$ such that $N \vDash \varphi(f, i, a_n)$.
Thus there exists $p \in G$ and hereditarily $\Fcal$-symmetric names $\dot{f}, \dot{i}$ and $\dot{a_n}$ for each of $f, i, a_n$ such that 
$$p \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_n})$$
By choosing $n$ and the names appropriately, we can find $\pi \in \Gcal$ such that the following conditions are additionally satisfied:
\begin{enumerate}
  \item $\pi \dot{f} = \dot{f}$
  \item $\pi \dot{i} = \dot{i}$
  \item $\pi n \ne n$
  \item There exists a hereditarily $\Fcal$-symmetric name $\dot{a_{\pi n}}$ of $a_{\pi n}$ such that $\pi \dot{a_n} = \dot{a_{\pi n}}$
  \item There exists $q \in G$ such that $q \leq_{\Pbb} p$ and $q \leq_{\Pbb} \pi p$
\end{enumerate}
By Lemma \ref{lem:symmetry_lemma}
$$\pi p \Vdash_{\text{HS}} \varphi(\pi \dot{f}, \pi \dot{i}, \pi \dot{a_n})$$
Thus 
$$\pi p \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_{\pi n}})$$
Therefore, by Lemma \ref{lem:strengthening}
$$q \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_n}) \text{ and } q \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_{\pi n}})$$
This means that $\Ncal \vDash \varphi(f, i, a_n)$ and $\Ncal \vDash \varphi(f, i, a_{\pi n})$, which implies that $f(i) = a_n$ and $f(i) = a_{\pi n}$.
Since $a_n$ and $a_{\pi n}$ are distinct, this is a contradiction. 
\end{proof}

\chapter{Isabelle/ZF and Formalization in Prior Work}
In this chapter, we introduce Isabelle/ZF, a proof assistant for ZF set theory, 
and the results from prior work used in the formalization of this study.

\section{Isabelle/ZF}
\section{Internalized First-Order Formulas}
\section{Forcing}

\chapter{Formalization of the Proof} 
\section{Introduction to the Formalization}
In this chapter, we

\section{Defining Symmetric Extensions}
\subsection{$\Pbb$-names}
\subsection{Automorphisms}
\subsection{Hereditarily Symmetric Names}
\subsection{Symmetric Extensions}

\section{Proving that Symmetric Extensions are Models of ZF}
\subsection{Relativized Forcing Relation $\Vdash_{\text{HS}}$}
\subsection{The Symmetric Lemma}
\subsection{Separation}
\subsection{Replacement}
\subsection{Other Axioms}

\section{Defining the Basic Cohen Model}
\subsection{The Notion of Forcing}
\subsection{The Group of Automorphisms}
\subsection{The Normal Filter}

@@

>>>>

\subsubsection{Our Approach}
The primary reason we chose Isabelle/ZF for this formalization is that Isabelle/ZF is a mature proof assistant for ZF set theory,
in particular, the formalization by Gunther et al. \cite{gunther_forcing} is a major advantage for this study.
Although there is also a formalization of forcing in Lean 3 by Han and van Doorn \cite{flypitch}, development for Lean 3 has already ended.

To use Gunther et al.'s formalization, we adopted the c.t.m. approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m. of ZF and constructed a model of ZF + $\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.
Our proof largely follows karagila's lecture note \cite{karagila}, with some parts also referencing Jech's books \cite{jech_set_theory,jech_AC}
as these resources align well with this approach.

The c.t.m. approach is a well-established method for relative consistency proofs in axiomatic set theory.
The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF + $\neg$AC is also consistent.
Assuming the consistency of ZF implies that a model of ZF exists, but strictly speaking, the existence of a c.t.m. cannot be derived from this assumption in ZF.
However, as explained in section \ref{sec:ctm}, we can prove the relative consistency of $\neg$AC with ZF if 
we can construct a model of ZF + $\neg$AC from a c.t.m. of ZF.
This kind of reasoning always arises in the c.t.m. approach, and ideally, 
we would like to formalize it as well, but this has not been achieved.

@@

\begin{comment}
  section2-...で説明するように、c.t.m.の存在はconzfからは導けないが、正当性が言えるのです
\end{comment}

@@

\chapter{Set-Theoretic Preliminaries}
In this chapter, we introduce the concepts of set theory used in the formalization of this study.
Our meta-theory is ZF, within which we explore ZF itself.

\section{ZF Set Theory and the Axiom of Choice}
We introduce the axioms of ZF set theory and the axiom of choice(AC).
We use first-order logic with the language of set theory, which consists only of only two relation symbols $\in$ and $=$. 
Formulas involving other mathmatical operators that may appear are considered abbreviations for formulas in this language.
Parentheses in formulas are omitted where no confusion arises.

\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\text{"} f \text{ is a function on } x \text{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF + AC is denoted by ZFC.
Additionally, we introduce the well-ordering theorem, as we treat AC in this form.

@@

\section{Forcing}
Forcing is a tequnique used in proving relative consistency and Independence.
We introduce basic concepts of forcing in the context of the c.t.m. approach.
In this approach, the relative consistency proof is achieved by using forcing to construct a extended model by adding new sets to an assumed c.t.m.
Let $M$ be a c.t.m. of ZF and $(\Pbb, \leq_{\Pbb})$ be a notion of forcing, which is a pre-ordered set in $M$ with a maximum element $1_{\Pbb}$.

\begin{dfn}
  We define $M^{\Pbb}$, the set of \emph{$\Pbb$-names}, by transfinite recursion on ordinals:
  \begin{enumerate}
    \item $M^{\Pbb}_0 = \emptyset$
    \item $M^{\Pbb}_{\alpha + 1} = \mathcal{P}^M(M^{\Pbb}_{\alpha} \times \Pbb)$
    \item $M^{\Pbb}_{\alpha} = \bigcup_{\beta < \alpha} M^{\Pbb}_{\beta}$ for a limit ordinal $\alpha$
    \item $M^{\Pbb} = \bigcup_{\alpha \in \text{Ord}} M^{\Pbb}_{\alpha}$
  \end{enumerate}
\end{dfn}
Where $\mathcal{P}^M$ denotes the power set operation in $M$. 
We often write a $\Pbb$-name with a dot, e.g., $\dot{x}$.
The least $\alpha$ such that $\dot{x} \in M^{\Pbb}_{\alpha + 1}$ is called the \emph{$\Pbb$-rank} of $\dot{x}$.
This allows us to define functions and relations by recursion on $\Pbb$-names.

@@

\begin{dfn} 
  Let $G$ be a generic filter on $\Pbb$ and $\dot{x} \in M^{\Pbb}$. We define the \emph{interpretation} of $\dot{x}$ denoted by $\dot{x}_G$ recursively with respect to the $\Pbb$-rank of $\dot{x}$:
  $$\dot{x}_G = \{\dot{y}_G \mid \exists p \in G (\langle \dot{y}, p \rangle \in \dot{x})\}$$ 
\end{dfn}

@@

By choosing $\Pbb$ appropriately, we can construct $M[G]$ with various properties.
What holds or does not hold in $M[G]$ can be identified using the forcing relation.
The forcing relation is defined recursively on formulas in a forcing language.
The forcing language is an extension of tha language of set theory by adding the elements of $M^{\Pbb}$ as constants.
\begin{dfn} %kunen
  We define the \emph{forcing relation} $\Vdash$ for formulas in the forcing language and $p \in \Pbb$ recursively
  \footnote{Specifically}
  as follows:
  \begin{enumerate}
    \item $p \Vdash \dot{x} = \dot{y} \Leftrightarrow \forall \dot{z} \in \text{dom}(\dot{x}) \cup \text{dom}(\dot{y})\,
\forall q \leq_{\Pbb} p \, (q \Vdash \dot{z} \in \dot{x} \leftrightarrow q \Vdash \dot{z} \in \dot{y})$
    \item $p \Vdash \dot{x} \in \dot{y} \Leftrightarrow \forall q \leq_{\Pbb} p\, \exists r \leq_{\Pbb} q \, 
    \exists s \in \Pbb \, \exists \dot{z} \in M^{\Pbb} \, (\langle \dot{z}, s \rangle \in \dot{y} \land r \leq_{\Pbb} s \land r \Vdash \dot{x} = \dot{z} )$
    \item $p \Vdash \phi \land \psi \Leftrightarrow p \Vdash \phi \land p \Vdash \psi$
    \item $p \Vdash \neg \phi \Leftrightarrow \forall q \leq_{\Pbb} p \, \neg (q \Vdash \phi)$
    \item $p \Vdash \exists x \phi(x) \Leftrightarrow \forall q \leq_{\Pbb} p \, \exists r \leq_{\Pbb} q \, \exists \dot{x} \in M^{\Pbb} \, (r \Vdash \phi(\dot{x}))$
  \end{enumerate}
\end{dfn}

@@

\section{The c.t.m. Approach}\label{sec:ctm}

@@

\begin{proof}
We outline the proof of this theorem as follows. 
For every $n \in \omega$, let $a_n$ be the following real number (a subset of $\omega$):
$$a_n = \{ m \in \omega \mid \exists p \in G (p(n, m) = 1) \}$$
Since $a_n$ are pairwise distinct, $A = \{ a_n \mid n \in \omega \}$ is an infinity set.
$A$ and every $a_n$ are elements of $\Ncal$.
$A$ serves as a counterexample to the well-ordering theorem in $\Ncal$.
Suppose for contradiction that $A$ is well-ordered in $\Ncal$, there exists a injection $f$ from $\omega$ to $A$ in $\Ncal$.
Let $\varphi(g,x,y)$ be a formula that represents the relation $g(x) = y$. 
For every $n \in \omega$ with $a_n \in \text{ran}(f)$, there exists $i \in \omega$ such that $N \vDash \varphi(f, i, a_n)$.
Thus there exists $p \in G$ and hereditarily $\Fcal$-symmetric names $\dot{f}, \dot{i}$ and $\dot{a_n}$ for each of $f, i, a_n$ such that 
$$p \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_n})$$
By choosing $n$ and the names appropriately, we can find a bijection $\pi$ on $\omega$ such that the following conditions are additionally satisfied:
\begin{enumerate}
  \item $\pi \dot{f} = \dot{f}$
  \item $\pi \dot{i} = \dot{i}$
  \item $\pi n \ne n$
  \item There exists a hereditarily $\Fcal$-symmetric name $\dot{a_{\pi n}}$ of $a_{\pi n}$ such that $\pi \dot{a_n} = \dot{a_{\pi n}}$
  \item There exists $q \in G$ such that $q \leq_{\Pbb} p$ and $q \leq_{\Pbb} \pi p$
\end{enumerate}
Note that some occurence of $\pi$ in above conditions refer to the induced automorphism on $\Pbb$ or $\Pbb$-names..
By Lemma \ref{lem:symmetry_lemma}
$$\pi p \Vdash_{\text{HS}} \varphi(\pi \dot{f}, \pi \dot{i}, \pi \dot{a_n})$$
Thus 
$$\pi p \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_{\pi n}})$$
Therefore, by Lemma \ref{lem:strengthening}
$$q \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_n}) \text{ and } q \Vdash_{\text{HS}} \varphi(\dot{f}, \dot{i}, \dot{a_{\pi n}})$$
This means that $\Ncal \vDash \varphi(f, i, a_n)$ and $\Ncal \vDash \varphi(f, i, a_{\pi n})$, which implies that $f(i) = a_n$ and $f(i) = a_{\pi n}$.
Since $a_n$ and $a_{\pi n}$ are distinct, this is a contradiction. 
\end{proof}

\chapter{Isabelle/ZF and Formalization in Prior Work}
In this chapter, we introduce Isabelle/ZF, a proof assistant for ZF set theory, 
and the results from prior work used in the formalization of this study.
\begin{comment}  
\section{Isabelle/ZF}
\section{Internalized First-Order Formulas}
\section{Forcing}  
\end{comment}

\chapter{Formalization of the Proof} 
\section{Defining Symmetric Extensions}
\begin{comment}
  \subsection{$\Pbb$-names}
\subsection{Automorphisms}
\subsection{Hereditarily Symmetric Names}
\subsection{Symmetric Extensions}
\end{comment}

\section{Proving that Symmetric Extensions are Models of ZF}
\begin{comment}
\subsection{Relativized Forcing Relation $\Vdash_{\text{HS}}$}
\subsection{The Symmetric Lemma}
\subsection{Separation}
\subsection{Replacement}
\subsection{Other Axioms}
\end{comment}

\section{Defining the Basic Cohen Model}
\begin{comment}
\subsection{The Notion of Forcing}
\subsection{The Group of Automorphisms}
\subsection{The Normal Filter}
\end{comment}

@@

\subsubsection{Acknowledgements}
